---
title: "Introduction to epidesc"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to epidesc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

The R package `epidesc` provides the tools to easily compute a series of epidemiological indicators to characterise different transmission profiles of infectious diseases. The work is based on the publication [How heterogeneous is the dengue transmission profile in Brazil? A study in six Brazilian states](https://doi.org/10.1371/journal.pntd.0010746) published in PLOS neglected tropical diseases by Iasmim Ferreira de Almeida, Raquel Martins Lana and Cláudia Torres Codeço in 2022.

While we include all the indicators contained in the original publication, `epidesc` aims to expand this list of indicators, welcoming contributions from researchers (please open and issue and we'll discuss!).

# Usage

To show the functionality of the package, we load weekly dengue data for municipalities in Rio de Janeiro.

```{r setup}
library(epidesc)
library(knitr)

data("dengueRio")
head(dengueRio)
```

## Data requirements

`epidesc` has a few simple data requirements:

1.  The input data has to be a data.frame with the cases and spatiotemporal identifiers. In our sample data, these would be the columns *casos* (dengue cases), *date* (temporal identifier), and *regional_geocode* (spatial identifier).
2.  The data needs to be weekly with dates in `Date` format.
3.  The time series needs to be regular, i.e. 7 days between consecutive observations.
4.  If an incidence descriptor is required, the population at risk also needs to be provided.

Don't worry though, if any of these aren't met, the package function will throw error messages to indicate to you what's happening!

## Formatting dates

The only required pre-processing step is to convert the dates into epidemiological yearweek format: yyyyww. We can easily do so by using the function `epiyearweek` as follows. Note that Brazilian epidemiological weeks start on a Sunday:

```{r}
dengue <- dengueRio
dengue$yearweek <- epiyearweek(dengue$date, start = "Sunday")
head(dengue)
```

## Descriptor list

To see the list of available descriptors and their required parameters, we can use the function `desc_list()`:

```{r, eval = FALSE}
desc_list()
```

In this table we can see the class of each indicator, its function name, description, and required parameters. This is information we will need when we specify which descriptors we want to compute.

## Computing descriptors

The first step when computing the descriptors is to decide which of them to compute and the values of the parameters. In out case, we are interested in computing the following indicators:

-   *Ap*
-   *Cmax* with 'x' equal to 3 and 6
-   *Inc* for 100,000 persons

To compute these indicators, we need to pass a nested list to the function `desc_year`. In the first level we state the output names of the descriptors that we want to use, and in the second level we specify the name of the function `fun` as in the table above as well as the parameters (`x`, `n` or `p` depending on the descriptor):

```{r}
descriptors <- list(
  Ap = list(fun = "Ap"),
  Cmax3 = list(fun = "Cmax", x = 3),
  Cmax6 = list(fun = "Cmax", x = 6),
  Inc = list(fun = "Inc", p = 100000)
)
```

Now we are almost ready to compute the descriptors. The last detail we need to consider is the epidemiological week we want to use to compute the yearly descriptors. For many diseases such as dengue in Brasil, the start of the season does not coincide with the 1st of January but much later in the year, so it makes much more sense to use this time point. We can specify the starting epidemiological week with the argument `sweek`. In our example we set it to week 41, meaning that we will compute yearly descriptors for year `t` starting from week 41 of year `t` until week 40 of year `t+1`.

Now we are ready to finally compute our descriptors with a call to `desc_year`:

```{r}
res <- desc_year(
  data = dengue, # Input data frame
  cases = "cases", # Name of the column of the cases
  time = "yearweek", # Name of the column with the epiyearweek
  space = "muni_code", # Name of the column with the spatial ID
  pop = "pop", # Column with population, needed because of 'Inc'
  sweek = 41, # Starting epiweek, 41 for the season of dengue in Brasil
  descriptors = descriptors # Descriptor list
)
```

Our descriptors have been computed! Let's have a look at the results of epidemiological year 2020 (from W41-2020 to W40-2021):

```{r}
head(res[res$epiyear == 2020, ])
```

Note that for epiyears where not all weeks are available or there are missing cases, the result will be NA, e.g.:

```{r}
head(res[res$epiyear == 2022, ])
```
