---
title: "Introduction to epidesc"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to epidesc}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

The R package `epidesc` provides the tools to easily compute a series of epidemiological indicators to characterise different transmission profiles of infectious diseases. The work is based on the publication [How heterogeneous is the dengue transmission profile in Brazil? A study in six Brazilian states](https://doi.org/10.1371/journal.pntd.0010746) published in PLOS neglected tropical diseases by Iasmim Ferreira de Almeida, Raquel Martins Lana and Cláudia Torres Codeço in 2022.

While we include all the indicators contained in the original publication, `epidesc` aims to expand this list of indicators, welcoming contributions from researchers (please open an issue and we'll discuss!).

# Usage

To show the functionality of the package, we load weekly dengue data for municipalities in Rio de Janeiro.

```{r setup}
library(epidesc)
library(knitr)

head(dengueRio)
```

## Data requirements

`epidesc` has a few simple data requirements:

1.  The input data has to be a data.frame with the cases and spatiotemporal identifiers. In our sample data, these would be the columns *casos* (dengue cases), *date* (temporal identifier), and *regional_geocode* (spatial identifier).
2.  The data needs to be weekly with dates in `Date` format.
3.  The time series needs to be regular, i.e. 7 days between consecutive observations.
4.  If an incidence descriptor is required, the population at risk also needs to be provided.

Don't worry though, if any of these aren't met, the package function will throw error messages to indicate to you what's happening!

## Formatting dates

The only required pre-processing step is to convert the dates into epidemiological yearweek format: yyyyww. We can easily do so by using the function `epiyearweek` as follows. Note that Brazilian epidemiological weeks start on a Sunday:
1. The input must be a `data.frame` containing the case counts and the corresponding spatial and temporal identifiers. In the sample dataset, these are the columns `cases` (number of dengue cases), `date` (temporal identifier), and `regional_geocode` (spatial identifier).
2. The data must be aggregated at the weekly level, with dates stored in `Date` format.
3. Each time series must be regular, with a 7-day interval between consecutive observations. If no cases are reported for a given spatial unit in a particular epidemiological week, that observation must still be included in the dataset with the case count set to zero. MAYBE PUT HERE HOW TO DO IT USING EXPAND.GRID???????? OR JUST SUGGESTING THE USER TO LOOK FOR HOW TO EXPAND A GRID????
4. If an incidence descriptor is required, the population at risk also needs to be provided.

Don't worry though, if any of these aren't met, the package function will throw error messages to let you know what's happening! 

## Formatting dates

The only required pre-processing step is to convert the dates into epidemiological yearweek format: yyyyww. We can easily do so by using the function `epiyearweek` as follows. 

Note that `epiyearweek` lets you choose whether epidemiological weeks start on a Sunday or a Monday via the `start` argument. If none is provided, the default is Sunday. Brazilian epidemiological weeks start on a Sunday, so this is the setting we'll use here.
Do not worry though, if any of these are not met, the package function will throw error messages to indicate to you what is happening! 

## Formatting dates

The only required pre-processing step is to convert the dates into epidemiological *yearweek* format: yyyyww. We can easily do so by using the function `epiyearweek` as follows. Here, we aggregate the data by the first day of each epidemiological week. 
Note that Brazilian epidemiological weeks start on a Sunday:

```{r}
dengueRio$yearweek <- epiyearweek(dengueRio$date, start = "Sunday")
head(dengueRio)
```

If you would like to check the start and end dates of the epidemiological weeks in the example data, you can access: [https://portalsinan.saude.gov.br/calendario-epidemiologico-2017](https://portalsinan.saude.gov.br/calendario-epidemiologico-2017). 

## Descriptor list

To see the list of available descriptors and their required parameters, we can use the function `desc_list()`:

I UPDATED FEW OF THEM IN THE README TABLE AND desc_list, NOT SURE IF IT IS WORKED WELL IN THE desc_lis.

```{r, eval = FALSE}
desc_list()
```

In this table we can see the class of each indicator, its function name, description, and required parameters. This is information we will need when we specify which descriptors we want to compute.

```{r, echo=FALSE}
kable(desc_list())
```

In this table we can see each indicator's class, function name, description, and required parameters. We will need this is information when we specify which descriptors we want to compute.

## Computing descriptors

The first step in computing the descriptors is to decide which ones to compute and what values to use for the parameters. In our case, we are interested in computing the following indicators:

* *Ap*
* *Cmax* with 'x' equal to 5
* *Cnf* with 'n' equal to 3 and 'x' equal to 5
* *Inc* for 100,000 persons

To compute these indicators, we need to pass a nested list to the `desc_year` function. In the first level, we state the output names of the descriptors that we want to use, and in the second level, we specify the name of the function `fun` as in the table above as well as the parameters (`x`, `n` or `p` depending on the descriptor): 



```{r}
descriptors <- list(
  Ap = list(fun = "Ap"),
  Cmax = list(fun = "Cmax", x = 5),
  Cnf3 = list(fun = "Cnf", n = 3, x = 5),
  Inc = list(fun = "Inc", p = 100000)
)
```

Now we are almost ready to compute the descriptors. The last detail we need to consider is the epidemiological week we want to use as the start of the year for computing yearly descriptors. For many diseases, such as dengue in Brazil, the start of the season does not coincide with the 1st of January but rather it occurs much later in the year, so it makes more sense to use this time point. We can specify the first epidemiological week with the argument `sweek`. In our example, we set it to week 41, meaning that yearly descriptors for year `t` will be computed starting from week 41 of year `t` until week 40 of year `t+1`.

Now we are ready to finally compute our descriptors with a call to `desc_year`:

FOR THE EPIYEAR, I THINK IT WOULD BE BETTER TO CALL IS LIKE A PERIOD: 2017/2018, 2018/2019 TO MAKE IT CLEAR THAT THE EPIYEAR HAS PART OF 2 YEARS. WOULD YOU REMEMBER WE'VE TALKED TO DO LIKE THIS? WHAT DO YOU THINK?

```{r}
res <- desc_year(
  data = dengueRio,            # Input data frame
  cases = "cases",          # Name of the column of the cases
  time = "yearweek",        # Name of the column with the epiyearweek
  space = "muni_code",      # Name of the column with the spatial ID
  pop = "pop",              # Column with population, needed because of 'Inc'
  sweek = 41,               # Starting epiweek, 41 for the dengue season in Brazil
  descriptors = descriptors # Descriptor list
)
```

AND ABOUT THE WEEK 53??? DO YOU REMEMBER WE TALKED ABOUT IT? TO ADD THE POSSIBILITY TO DEAL WITH IT, LIKE SPLIT THIS WEEK? WE ALSO DID A FUNCTION FOR IT. I THINK IN THE OLD REPO, WE HAVE IT. 

Our descriptors have been computed! Let's have a look at the results of epidemiological year 2020 (from W41-2020 to W40-2021):

```{r}
head(res[res$epiyear == 2020, ])
```

Note that for epiyears where not all weeks are available or there are missing cases, the result will be NA, e.g.:
I THINK IT IS NOT ALLOWED TO HAVE MISSING CASES. MAYBE WE HAVE TO BE CLEAR ABOUT IT. I TRIED ABOVE IN THE REQUIREMENTS. MAYBE EXPLAIN THE BORDERS OF THE TIME SERIES WILL BE NOT USED IN THE CALCULATION IF YOU CHOOSE TO CHANGE FROM CALENDAR YEAR TO EPI YEAR.  

```{r}
head(res[res$epiyear == 2022, ])
```
